Refactoring `nls()`: a constructive critique

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
add code structure

>>>>

## Issue: code structure and documentation for maintenance

The `nls()` code is structured in a way that inhibits both maintenance and improvement.
In particular, the iterative setup is such that introduction of Marquardt stabilization
is not easily available. To obtain performance when the code was written, a large portion
was written in C with consequent calls and returns that
severely complicate the code. Over time, R has become much more efficient on modern computers, and
the need to use compiled C and Fortran is less critical. Moreover, the burden for maintenance
could be reduced by moving code entirely to R.

While R and its packages are generally very well documented for usage, that documentation
rarely extends as far as it might for code maintenance. 
The paucity of such documentation is exacerbated by the mixed R/C/Fortran code base,
where seemingly trivial differences in things like representation of numbers or
base index for arrays lead to traps for unwary programmers.

<!-- For `nls()` the concern is demonstrated by a short email query from John Nash -->
<!-- to Doug Bates and his reply. -->
<!-- This is NOT a criticism of Prof. Bates' (or any other) work, but a reflection on  -->
<!-- how difficult it is to develop code in this subject area and to keep it maintainable.  -->
<!-- We have experienced similar loss of understanding for some of our own codes. -->
<!--     ... -->
<!--     https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R -->
<!--     ... has the test problem and the expanded code. Around line 367 is where we are -->
<!--     scratching our heads. The function code (from nlsModel()) is in the commented  -->
<!--     lines below the call. This is -->
<!--           # > setPars -->
<!--           # function(newPars) { -->
<!--           #   setPars(newPars) -->
<!--           #   resid <<- .swts * (lhs - (rhs <<- getRHS())) # envir = thisEnv {2 x} -->
<!--           #   dev   <<- sum(resid^2) # envir = thisEnv -->
<!--           #   if(length(gr <- attr(rhs, "gradient")) == 1L) gr <- c(gr) -->
<!--           #   QR <<- qr(.swts * gr) # envir = thisEnv -->
<!--           #   (QR$rank < min(dim(QR$qr))) # to catch the singular gradient matrix -->
<!--           # } -->
<!--     I'm anticipating that we will be able to set up a (possibly inefficient) code -->
<!--     with documentation that will be easier to follow and test, then gradually figure -->
<!--     out how to make it more efficient. -->
<!--     The equivalent from minpack.lm is -->
<!--     setPars = function(newPars) { -->
<!--                 setPars(newPars) -->
<!--                 assign("resid", .swts * (lhs - assign("rhs", getRHS(), -->
<!--                     envir = thisEnv)), envir = thisEnv) -->
<!--                 assign("dev", sum(resid^2), envir = thisEnv) -->
<!--                 assign("QR", qr(.swts * attr(rhs, "gradient")), envir = thisEnv) -->
<!--                 return(QR$rank < min(dim(QR$qr))) -->
<!--             } -->
<!--     In both there is the recursive call, which must have a purpose I don't understand. -->
<!-- The reply was -->
<!--     I'm afraid that I don't know the purpose of the recursive call either.   -->
<!--     I know that I wrote the code to use a closure for the response, covariates, etc.,  -->
<!--     but I don't recall anything like a recursive call being necessary. -->
<!--      I don't think I will be of much help.  My R skills have atrophied to the point  --> 
<!--      where I wouldn't even know how to start exploring what is happening in the first  -->
<!--      call as opposed to the recursive call. --> 
<!--  ``` -->

# Goals and progress of our effort to improve `nls()`
